---
title: "Diagnostic distribution plots with plotness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostic distribution plots with plotness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
set.seed(20210)
```

## What these plots are for

When you have count data, a natural first question is *which discrete
distribution does it look like* — Poisson, binomial, or negative binomial?
**Distribution-ness plots** (Hoaglin & Tukey) answer that visually. Each one
transforms the observed frequencies through a distribution-specific *metameter*
chosen so that, **if the data really follow that distribution, the points fall
on a straight line.** Systematic curvature or scatter away from the line is the
diagnostic signal that the distribution does not fit.

`plotness()` builds these plots with `ggplot2`. It is adapted from
`vcd::distplot()`, with two additions: the output is a ggplot you can theme and
extend, and `plot = FALSE` hands you the underlying data so you can draw the
plot however you like.

```{r}
library(plotness)
library(ggplot2)
```

## Poissonness plot

The straight-line case. We simulate Poisson counts and check them against the
Poisson metameter — the points should hug the fitted "perfect distribution"
line, and the on-plot panel reports the slope, intercept, and the implied
parameter estimate (`exp(slope)` ≈ the rate λ).

```{r poisson}
pois_counts <- rpois(200, lambda = 6)
plotness(pois_counts, type = "poisson")
```

## Binomialness plot

For binomial data you supply `size` (the number of trials). If you omit it,
`plotness()` falls back to the maximum observed count and warns you — fine for a
quick look, but pass the real value when you know it.

```{r binomial}
binom_counts <- rbinom(200, size = 10, prob = 0.3)
plotness(binom_counts, type = "binomial", size = 10)
```

## Negative binomialness plot

When `size` is left `NULL` for `type = "nbinomial"`, the dispersion parameter is
estimated from the data.

```{r nbinomial}
nbinom_counts <- rnbinom(200, size = 4, prob = 0.4)
plotness(nbinom_counts, type = "nbinomial")
```

## Diagnosing a *bad* fit

The plots earn their keep when the distribution is wrong. Negative-binomial data
is overdispersed relative to Poisson, so feeding it to the **Poissonness** plot
should bend the points away from the line:

```{r misfit}
overdispersed <- rnbinom(200, size = 1.5, prob = 0.2)
plotness(overdispersed, type = "poisson",
         title = "Overdispersed counts on a Poissonness plot")
```

The curvature — points sagging below the line in the middle — is the tell-tale
sign that a single Poisson rate does not describe these counts.

## Getting the data instead of the plot

Set `plot = FALSE` to receive the data frame that backs the plot. This lets you
use your own graphics stack, or pull out the fitted line and confidence
intervals for further analysis.

```{r data}
df <- plotness(pois_counts, type = "poisson", plot = FALSE)
head(df)
```

The columns are:

| Column | Meaning |
|---|---|
| `Counts` | the observed count value |
| `Freq` | how many times it occurred |
| `Metameter` | the observed point on the plot |
| `CI.center`, `CI.width`, `CI.lower`, `CI.upper` | confidence interval around the smoothed metameter |
| `y_line` | the fitted "perfect distribution" line at each count |

A minimal hand-rolled version of the plot from that data:

```{r data-plot}
ggplot(df, aes(Counts)) +
  geom_line(aes(y = y_line), colour = "blue") +
  geom_point(aes(y = Metameter), colour = "red") +
  labs(y = "Metameter", title = "Built from plotness(plot = FALSE)") +
  theme_bw()
```

## Useful options

A few arguments worth knowing:

* `conf_int = FALSE` drops the confidence-interval bars.
* `label = FALSE` removes the on-plot fit summary; `legend = FALSE` removes the
  legend.
* `title`, `xlab`, `ylab` override the text; `xlim` restricts the x-axis range.
* For `type = "poisson"`, supplying `lambda` produces a *leveled* Poissonness
  plot (useful when comparing against a hypothesized rate).

```{r options}
plotness(pois_counts, type = "poisson",
         conf_int = FALSE, label = FALSE,
         title = "Stripped-down Poissonness plot")
```

## Reference

Hoaglin, D. C. & Tukey, J. W. (1985). Checking the shape of discrete
distributions. In *Exploring Data Tables, Trends, and Shapes*. The underlying
machinery follows `vcd::distplot()` (Meyer, Zeileis & Hornik).
